Problem of Step 4 (debias) of fGFPCA on NHANES data

### 2024-01-31 Docs/NhanesApplExample.Rmd
reversed eigenvalue problems

set.seed(121)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
theme_set(theme_minimal())
```

@@

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
K <- 3
source(here("Code/OutsampBayes.R"))
```

Fit fGFPCA model

Because step 4 (debias) is very time-consuming, this example is fit on a subset of NHANES dataset, including 1000 participants as the training set

```{r data_subset}
# head(df)
df <- df %>% rename(id=SEQN, Y=Z)

Ntr <- 1000

train_id <- unique(df$id)[1:Ntr]
train_df <- df %>% filter(id %in% train_id)

# additional parameters
J <- length(unique(df$sind))
```

@@

train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
train_df$bin <- as.numeric(as.character(train_df$bin))
```

# Step 2:  Local GLMM

```{r}
# fit model on the training set
train_bin_lst <- split(train_df, f = train_df$bin)

t1=Sys.time()
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
t2= Sys.time()
# t2-t1 
```

```{r}
df_est_latent <- bind_rows(df_est_latent) 
# head(df_est_latent)

# example estimated latent function
rand_id <- sample(train_id, 4)
df_est_latent %>% 
  filter(id %in% rand_id) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
  geom_point(aes(x=sind, y = Y, group = id, col = "outcome"), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

# Step 3: FPCA

Follow up on the email yesterday, we would like to see what would happen if we only take the first **three** PC functions.

```{r}
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=Ntr, 
                         ncol=n_bin, byrow = F) 
# row index subject, column binned time

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
# keep 3 PCs
t2 <- Sys.time()
# t2-t1

# fpca_mod$evalues
```

```{r PC}
data.frame(t = mid, fpca_mod$efunctions) %>%
  rename(PC1 = X1, PC2=X2, PC3=X3) %>%
  pivot_longer(2:4, names_to = "PC") %>%
  ggplot()+
  geom_line(aes(x=t, y=value, col = PC))+
  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
```

- Eigenvalues: `r fpca_mod$evalues`
- Proportion of variance explained with four PCs: `r fpca_mod$pve`

# Step 4: Re-evaluation

## Extend the eigenfunctions back to the original grid with B-spline basis

```{r interplate}
# Re-evaluation
## grid extension
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

@@

df_phi <- matrix(NA, J, K) 
for(k in 1:K){
  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
  df_phi[,k] <- Bnew %*% coef(lm_mod)
}# project binned eigenfunctions onto the original grid
```

Below presents the result of re-evaluation. Points are values of PC functions on the binned grid estimated from FPCA, and lines are values of re-evaluated PC functions on the original measurement grid.

```{r}
df_pc1 <- data.frame(t=1:J,  df_phi) %>%
  rename(PC1=X1, PC2=X2, PC3=X3) %>%
  pivot_longer(2:4, names_to = "PC")

df_pc2 <- data.frame(t = mid, fpca_mod$efunctions) %>%
  rename(PC1=X1, PC2=X2, PC3=X3) %>%
  pivot_longer(2:4, names_to = "PC")

left_join(df_pc1, df_pc2, by = c("t", "PC")) %>%
  ggplot()+
  geom_line(aes(x=t, y=value.x, col = PC), linewidth = 0.2)+
  geom_point(aes(x=t, y=value.y, col = PC), na.rm = T, alpha = 0.5, size = 0.5)+
  labs(x="Time", y="", title = "Re-evaluate eigenfunctions on the original grid")

```

## Debias the eigenvalues with GLMM

Again, here we use the three PC functions in the random effect part.

```{r revaluation}
df_phi <- data.frame(sind = 1:J, df_phi)
colnames(df_phi) <- c("sind", paste0("phi", 1:K))
train_df <- train_df %>% 
  left_join(df_phi, by = "sind")
train_df$id <- as.factor(train_df$id)

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re"),
                     # s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
# t2-t1 # 3.5 minutes
```

It seems that the eigenvalues are still losing their decreasing nature. The eigenvalue corresponding to PC2 is still smaller than PC3. Specifically

- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:4]`
- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:4]`
- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:3]`

Also, the proportion of these two set of estimates is not constant, as below:

```{r, class.source = 'fold-show'}
# proportion of biased eigenvalues to re-evaluated eigenvalues
fpca_mod$evalues/(1/debias_glmm$sp[2:4])
```

>>>>

set.seed(130)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
library(arsenal)
theme_set(theme_minimal())
```

@@

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
K <- 4
source(here("Code/OutsampBayes.R"))
```

# Fit fGFPCA model

```{r data_subset}
# head(df)
df <- df %>% rename(id=SEQN, Y=Z)
N <- length(unique(df$id))
J <- length(unique(df$sind))
```

@@

df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
df$bin <- as.numeric(as.character(df$bin))
```

# Step 2-4:  Local GLMM, fPCA, splines interpolation and GLMM reevaluation

I'd like to investigate with different sample sizes.

```{r, cache=TRUE, results='hide'}
# sample size 
N_sub <- rep(500, 4)
fpca_model_list <- list()
interp_pc_est <- list()
debias_model_list <- list()
time_vec <- rep(NA, length(N_sub))

pb <- txtProgressBar(min=0, max = length(N_sub), style = 3)
for(i in seq_along(N_sub)){

  # training sample of a certain size
  Ntr <- N_sub[i]
  train_id <- sample(unique(df$id), size = Ntr) 
  train_df <- df %>% filter(id %in% train_id)

  t1=Sys.time()
  # local GLMM
  train_bin_lst <- split(train_df, f = train_df$bin)
  df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
  df_est_latent <- bind_rows(df_est_latent)

  # FPCA
  uni_eta_hat <- df_est_latent %>% filter(bin==sind)
  mat_est_unique <- matrix(uni_eta_hat$eta_hat, nrow=Ntr, ncol=n_bin, byrow = F) 
  fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
  fpca_model_list[[i]] <- fpca_mod

  # interpolation
  p <- 3 # order of b splines 
  knots <- 35 # number of knots (same from FPCA model)
  knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
  knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

@@

  df_phi <- matrix(NA, J, K) 
  for(k in 1:K){
    lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
    df_phi[,k] <- Bnew %*% coef(lm_mod)
  }# project binned eigenfunctions onto the original grid
  interp_pc_est[[i]] <- df_phi

  # debias model
  df_phi <- data.frame(sind = 1:J, df_phi)
  colnames(df_phi) <- c("sind", paste0("phi", 1:K))
  train_df <- train_df %>% 
    left_join(df_phi, by = "sind")
  train_df$id <- as.factor(train_df$id)
  debias_glmm <- bam(Y ~ s(sind, bs="cr")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
  debias_model_list[[i]] <- debias_glmm
  t2 <- Sys.time()

  time_vec[i] <- as.numeric(t2-t1, unit = "secs")

  setTxtProgressBar(pb, i)
}

close(pb)
```

<!-- - Time spent on model fitting -->

```{r, eval=FALSE}
time_vec
```

<!-- - FPCA eigenvalues -->

```{r, eval=FALSE}
lapply(fpca_model_list, function(x){x$evalues})
lapply(fpca_model_list, function(x){x$pve})
```

- Eigenvalues from debias models

```{r}
lapply(debias_model_list, function(x){1/x$sp[2:5]})
```

- PC functions

```{r}
pc_plot_list <- list()

for(i in  1:length(N_sub)){

  df_phi1 <- data.frame(t=mid, fpca_model_list[[i]]$efunctions)
  df_phi2 <- data.frame(t = 1:J, interp_pc_est[[i]])

  this_plot <- left_join(df_phi2, df_phi1, by = "t") %>%
    ggplot()+
    geom_point(aes(x=t, y=X1.y, col = "PC1"), na.rm = T, size = 0.5)+
    geom_point(aes(x=t, y=X2.y, col = "PC2"), na.rm = T, size = 0.5)+
    geom_point(aes(x=t, y=X3.y, col = "PC3"), na.rm = T, size = 0.5)+
    geom_point(aes(x=t, y=X4.y, col = "PC4"), na.rm = T, size = 0.5)+
    geom_line(aes(x=t, y=X1.x, col = "PC1"))+
    geom_line(aes(x=t, y=X2.x, col = "PC2"))+
    geom_line(aes(x=t, y=X3.x, col = "PC3"))+
    geom_line(aes(x=t, y=X4.x, col = "PC4"))+
    labs(x="time", y = "", title = paste0("Iter", i))

  pc_plot_list[[i]] <- this_plot

}

ggarrange(plotlist = pc_plot_list, common.legend = T)
```

The "reverse eigenvalue problem" does not always happen. When training sample size is small (like maybe 200), the problem happened less. So this is definately a numeric problem. But a very small sample also would induce sigularity issues.

Some other thoughts:

1. What about estimation method of smooth parameter? It seems GCV doesn't treat model as mixed effects. REML may be too slow without the "discrete=T" option on the full dataset. 
2. Is is possible to impose monotone constraint on smoothing parameter estimation?
